# Book contents {-}

```{r setup, echo=FALSE, results="asis"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE)
library(BiocStyle)
```

The OSCA book is actually a collection of several sub-books spanning a variety of topics and different levels of assumed reader knowledge.
Each sub-book is generated using `r CRANpkg("bookdown")`, compiled twice a week to ensure that the examples still run on the current R/Bioconductor code base.
Interlinking between related topics enable readers to seamlessly transition within a sub-book, and even between different sub-books.
To get started, click on the links below to navigate to each sub-book.

## `r Biocbook("OSCA.intro", label="Introduction")` {-}

```{r, results="asis", echo=FALSE}
text <- packageDescription("OSCA.intro")$Description
remainders <- strsplit(text, "\\.\\s")[[1]][-1]
remainders <- sub("^\\s+", "", remainders)
remainders <- gsub("SingleCellExperiment", "`SingleCellExperiment`", remainders)
remainders <- paste(remainders, collapse=". ")
cat(remainders)
```

## `r Biocbook("OSCA.basic", label="Basics")` {-}

```{r, results="asis", echo=FALSE}
text <- packageDescription("OSCA.basic")$Description
remainders <- strsplit(text, "\\.\\s")[[1]][-1]
remainders <- sub("^\\s+", "", remainders)
remainders <- paste(remainders, collapse=". ")
cat(remainders, sep="\n")
```

## `r Biocbook("OSCA.advanced", label="Advanced")` {-}

```{r, results="asis", echo=FALSE}
text <- packageDescription("OSCA.advanced")$Description
remainders <- strsplit(text, "\\.\\s")[[1]][-1]
remainders <- sub("^\\s+", "", remainders)
remainders <- paste(remainders, collapse=". ")
cat(remainders, sep="\n")
```

## `r Biocbook("OSCA.workflow", label="Workflows")` {-}

```{r, results="asis", echo=FALSE}
text <- packageDescription("OSCA.workflows")$Description
remainders <- strsplit(text, "\\.\\s")[[1]][-1]
remainders <- gsub("SingleCellExperiment", "`SingleCellExperiment`", remainders)
remainders <- sub("^\\s+", "", remainders)
remainders <- paste(remainders, collapse=". ")
cat(remainders, sep="\n")
```
