# Nestorowa mouse HSC (Smart-seq2) 

```{r unref-setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble(cache=TRUE)
```

## Introduction

This performs an analysis of the mouse haematopoietic stem cell (HSC) dataset generated with Smart-seq2 [@nestorowa2016singlecell].

## Data loading

```{r data-loading}
library(scRNAseq)
sce.nest <- NestorowaHSCData()
```

```{r gene-annotation}
library(AnnotationHub)
ens.mm.v97 <- AnnotationHub()[["AH73905"]]
anno <- select(ens.mm.v97, keys=rownames(sce.nest), 
    keytype="GENEID", columns=c("SYMBOL", "SEQNAME"))
rowData(sce.nest) <- anno[match(rownames(sce.nest), anno$GENEID),]
```

After loading and annotation, we inspect the resulting `SingleCellExperiment` object:

```{r}
sce.nest
```

## Quality control

```{r}
unfiltered <- sce.nest
```

For some reason, no mitochondrial transcripts are available, so we will perform quality control using the spike-in proportions only.

```{r quality-control-grun}
library(scater)
stats <- perCellQCMetrics(sce.nest)
qc <- quickPerCellQC(stats, percent_subsets="altexps_ERCC_percent")
sce.nest <- sce.nest[,!qc$discard]
```

We examine the number of cells discarded for each reason.

```{r}
colSums(as.matrix(qc))
```

We create some diagnostic plots for each metric (Figure \@ref(fig:unref-nest-qc-dist)).

```{r unref-nest-qc-dist, fig.wide=TRUE, fig.cap="Distribution of each QC metric across cells in the Nestorowa HSC dataset. Each point represents a cell and is colored according to whether that cell was discarded."}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- qc$discard

gridExtra::grid.arrange(
    plotColData(unfiltered, y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, y="altexps_ERCC_percent",
        colour_by="discard") + ggtitle("ERCC percent"),
    ncol=2
)
```

## Normalization

```{r normalization}
library(scran)
set.seed(101000110)
clusters <- quickCluster(sce.nest)
sce.nest <- computeSumFactors(sce.nest, clusters=clusters)
sce.nest <- logNormCounts(sce.nest)
```

We examine some key metrics for the distribution of size factors, and compare it to the library sizes as a sanity check (Figure \@ref(fig:unref-nest-norm)).

```{r}
summary(sizeFactors(sce.nest))
```

```{r unref-nest-norm, fig.cap="Relationship between the library size factors and the deconvolution size factors in the Nestorowa HSC dataset."}
plot(librarySizeFactors(sce.nest), sizeFactors(sce.nest), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", log="xy")
```

## Variance modelling

We use the spike-in transcripts to model the technical noise as a function of the mean (Figure \@ref(fig:unref-nest-var)).

```{r variance-modelling}
set.seed(00010101)
dec.nest <- modelGeneVarWithSpikes(sce.nest, "ERCC")
top.nest <- getTopHVGs(dec.nest, prop=0.1)
```

```{r unref-nest-var, fig.cap="Per-gene variance as a function of the mean for the log-expression values in the Nestorowa HSC dataset. Each point represents a gene (black) with the mean-variance trend (blue) fitted to the spike-ins (red)."}
plot(dec.nest$mean, dec.nest$total, pch=16, cex=0.5,
    xlab="Mean of log-expression", ylab="Variance of log-expression")
curfit <- metadata(dec.nest)
curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
points(curfit$mean, curfit$var, col="red")
```

## Dimensionality reduction

```{r dimensionality-reduction}
set.seed(101010011)
sce.nest <- denoisePCA(sce.nest, technical=dec.nest, subset.row=top.nest)
sce.nest <- runTSNE(sce.nest, dimred="PCA")
```

We check that the number of retained PCs is sensible.

```{r}
ncol(reducedDim(sce.nest, "PCA"))
```

## Clustering

```{r clustering}
snn.gr <- buildSNNGraph(sce.nest, use.dimred="PCA")
colLabels(sce.nest) <- factor(igraph::cluster_walktrap(snn.gr)$membership)
```

```{r}
table(colLabels(sce.nest))
```

```{r unref-nest-tsne, fig.cap="Obligatory $t$-SNE plot of the Nestorowa HSC dataset, where each point represents a cell and is colored according to the assigned cluster."}
plotTSNE(sce.nest, colour_by="label")
```

## Marker gene detection

```{r marker-detection}
markers <- findMarkers(sce.nest, colLabels(sce.nest), 
    test.type="wilcox", direction="up", lfc=0.5,
    row.data=rowData(sce.nest)[,"SYMBOL",drop=FALSE])
```

```{r, echo=FALSE}
cluster <- "8"
```

To illustrate the manual annotation process, we examine the marker genes for one of the clusters.
Upregulation of _Car2_, _Hebp1_ amd hemoglobins indicates that cluster `r cluster` contains erythroid precursors.

```{r unref-heat-nest-markers, fig.asp=1.5, fig.cap=sprintf("Heatmap of the AUCs for the top marker genes in cluster %s compared to all other clusters.", cluster)}
chosen <- markers[['8']]
best <- chosen[chosen$Top <= 10,]
aucs <- getMarkerEffects(best, prefix="AUC")
rownames(aucs) <- best$SYMBOL

library(pheatmap)
pheatmap(aucs, color=viridis::plasma(100))
```

```{r, echo=FALSE}
# Checking my statements:
stopifnot(identical(chosen, markers[[cluster]]))
stopifnot(all(c("Car2", "Hba-a2", "Hba-a1", "Hebp1") %in% rownames(aucs)))
```

## Cell type annotation

```{r cell-annotation}
library(SingleR)
mm.ref <- MouseRNAseqData()

# Renaming to symbols to match with reference row names.
renamed <- sce.nest
rownames(renamed) <- uniquifyFeatureNames(rownames(renamed),
    rowData(sce.nest)$SYMBOL)
labels <- SingleR(renamed, mm.ref, labels=mm.ref$label.fine)
```

Most clusters are not assigned to any single lineage (Figure \@ref(fig:unref-assignments-nest)), which is perhaps unsurprising given that HSCs are quite different from their terminal fates.
Cluster `r cluster` is considered to contain erythrocytes, which is roughly consistent with our conclusions from the marker gene analysis above.

```{r unref-assignments-nest, fig.cap="Heatmap of the distribution of cells for each cluster in the Nestorowa HSC dataset, based on their assignment to each label in the mouse RNA-seq references from the _SingleR_ package."}
tab <- table(labels$labels, colLabels(sce.nest))
pheatmap(log10(tab+10), color=viridis::viridis(100))
```

```{r, echo=FALSE}
test <- tab[,"8"]
stopifnot(test["Erythrocytes"]/sum(test) > 0.9)
```

## Miscellaneous analyses

This dataset also contains information about the protein abundances in each cell from FACS.
There is barely any heterogeneity in the chosen markers across the clusters (Figure \@ref(fig:unref-nest-facs));
this is perhaps unsurprising given that all cells should be HSCs of some sort.

```{r unref-nest-facs, fig.cap="Heatmap of the centered log-average intensity for each target protein quantified by FACS in the Nestorowa HSC dataset."}
Y <- colData(sce.nest)$FACS
keep <- rowSums(is.na(Y))==0 # Removing NA intensities.

se.averaged <- sumCountsAcrossCells(t(Y[keep,]), 
    colLabels(sce.nest)[keep], average=TRUE)
averaged <- assay(se.averaged)

log.intensities <- log2(averaged+1)
centered <- log.intensities - rowMeans(log.intensities)
pheatmap(centered, breaks=seq(-1, 1, length.out=101))
```

```{r, echo=FALSE}
# Check that my trash talk is correct.
rr <- rowRanges(centered)
stopifnot(all(rr[,2] - rr[,1] < 1))
```

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
