# Human PBMC with surface proteins (10X Genomics)

```{r unref-setup, cache=FALSE, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble(cache=TRUE)
```

## Introduction

Here, we describe a brief analysis of _yet another_ peripheral blood mononuclear cell (PBMC) dataset from 10X Genomics [@zheng2017massively].
Data are publicly available from the [10X Genomics website](https://support.10xgenomics.com/single-cell-vdj/datasets/3.0.0/vdj_v1_mm_c57bl6_pbmc_5gex), from which we download the filtered gene/barcode count matrices for gene expression and cell surface proteins.
Note that most of the repertoire-related steps will be discussed in Chapter \@ref(repertoire-seq), this workflow mostly provides the baseline analysis for the expression data.

## Data loading

```{r loading}
library(BiocFileCache)
bfc <- BiocFileCache(ask=FALSE)
exprs.data <- bfcrpath(bfc, file.path(
    "http://cf.10xgenomics.com/samples/cell-vdj/3.1.0",
    "vdj_v1_hs_pbmc3",
    "vdj_v1_hs_pbmc3_filtered_feature_bc_matrix.tar.gz"))
untar(exprs.data, exdir=tempdir())

library(DropletUtils)
sce.pbmc <- read10xCounts(file.path(tempdir(), "filtered_feature_bc_matrix"))
sce.pbmc <- splitAltExps(sce.pbmc, rowData(sce.pbmc)$Type)
```

## Quality control

```{r}
unfiltered <- sce.pbmc
```

We discard cells with high mitochondrial proportions and few detectable ADT counts.

```{r quality-control}
library(scater)
is.mito <- grep("^MT-", rowData(sce.pbmc)$Symbol)
stats <- perCellQCMetrics(sce.pbmc, subsets=list(Mito=is.mito))

high.mito <- isOutlier(stats$subsets_Mito_percent, type="higher")
low.adt <- stats$`altexps_Antibody Capture_detected` < nrow(altExp(sce.pbmc))/2

discard <- high.mito | low.adt
sce.pbmc <- sce.pbmc[,!discard]
```

We examine some of the statistics:

```{r}
summary(high.mito)
summary(low.adt)
summary(discard)
```

We examine the distribution of each QC metric (Figure \@ref(fig:unref-pbmc-adt-qc)).

```{r unref-pbmc-adt-qc, fig.wide=TRUE, fig.cap="Distribution of each QC metric in the PBMC dataset, where each point is a cell and is colored by whether or not it was discarded by the outlier-based QC approach."}
colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- discard

gridExtra::grid.arrange(
    plotColData(unfiltered, y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, y="subsets_Mito_percent",
        colour_by="discard") + ggtitle("Mito percent"),
    plotColData(unfiltered, y="altexps_Antibody Capture_detected",
        colour_by="discard") + ggtitle("ADT detected"),
    ncol=2
)
```

We also plot the mitochondrial proportion against the total count for each cell, as one does (Figure \@ref(fig:unref-pbmc-adt-qc-mito)).

```{r unref-pbmc-adt-qc-mito, fig.cap="Percentage of UMIs mapped to mitochondrial genes against the totalcount for each cell."}
plotColData(unfiltered, x="sum", y="subsets_Mito_percent",
    colour_by="discard") + scale_x_log10()
```

## Normalization

Computing size factors for the gene expression and ADT counts.

```{r normalization}
library(scran)

set.seed(1000)
clusters <- quickCluster(sce.pbmc)
sce.pbmc <- computeSumFactors(sce.pbmc, cluster=clusters)
altExp(sce.pbmc) <- computeMedianFactors(altExp(sce.pbmc))
sce.pbmc <- logNormCounts(sce.pbmc, use_altexps=TRUE)
```

We generate some summary statistics for both sets of size factors:

```{r}
summary(sizeFactors(sce.pbmc))
summary(sizeFactors(altExp(sce.pbmc)))
```

We also look at the distribution of size factors compared to the library size for each set of features (Figure \@ref(fig:unref-norm-pbmc-adt)).

```{r unref-norm-pbmc-adt, fig.wide=TRUE, fig.cap="Plot of the deconvolution size factors for the gene expression values (left) or the median-based size factors for the ADT expression values (right) compared to the library size-derived factors for the corresponding set of features. Each point represents a cell."}
par(mfrow=c(1,2))
plot(librarySizeFactors(sce.pbmc), sizeFactors(sce.pbmc), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", 
    main="Gene expression", log="xy")
plot(librarySizeFactors(altExp(sce.pbmc)), sizeFactors(altExp(sce.pbmc)), pch=16,
    xlab="Library size factors", ylab="Median-based factors", 
    main="Antibody capture", log="xy")
```

## Dimensionality reduction

We omit the PCA step for the ADT expression matrix, given that it is already so low-dimensional,
and progress directly to $t$-SNE and UMAP visualizations.

```{r dimensionality-reduction}
set.seed(100000)
altExp(sce.pbmc) <- runTSNE(altExp(sce.pbmc))

set.seed(1000000)
altExp(sce.pbmc) <- runUMAP(altExp(sce.pbmc))
```

## Clustering

We perform graph-based clustering on the ADT data and use the assignments as the column labels of the alternative Experiment.

```{r clustering}
g.adt <- buildSNNGraph(altExp(sce.pbmc), k=10, d=NA)
clust.adt <- igraph::cluster_walktrap(g.adt)$membership
colLabels(altExp(sce.pbmc)) <- factor(clust.adt)
```

We examine some basic statistics about the size of each cluster, 
their separation (Figure \@ref(fig:unref-clustmod-pbmc-adt))
and their distribution in our $t$-SNE plot (Figure \@ref(fig:unref-tsne-pbmc-adt)).

```{r}
table(colLabels(altExp(sce.pbmc)))
```

```{r unref-clustmod-pbmc-adt, fig.cap="Heatmap of the pairwise cluster modularity scores in the PBMC dataset, computed based on the shared nearest neighbor graph derived from the ADT expression values."}
library(bluster)
mod <- pairwiseModularity(g.adt, clust.adt, as.ratio=TRUE)

library(pheatmap)
pheatmap::pheatmap(log10(mod + 10), cluster_row=FALSE, cluster_col=FALSE,
    color=colorRampPalette(c("white", "blue"))(101))
```

```{r unref-tsne-pbmc-adt, fig.cap="Obligatory $t$-SNE plot of PBMC dataset based on its ADT expression values, where each point is a cell and is colored by the cluster of origin. Cluster labels are also overlaid at the median coordinates across all cells in the cluster."}
plotTSNE(altExp(sce.pbmc), colour_by="label", text_by="label", text_col="red")
```

We perform some additional subclustering using the expression data to mimic an _in silico_ FACS experiment.

```{r subclustering}
set.seed(1010010)
subclusters <- quickSubCluster(sce.pbmc, clust.adt,
    prepFUN=function(x) {
        dec <- modelGeneVarByPoisson(x)
        top <- getTopHVGs(dec, prop=0.1)
        denoisePCA(x, dec, subset.row=top)
    }, 
    clusterFUN=function(x) {
        g.gene <- buildSNNGraph(x, k=10, use.dimred = 'PCA')
        igraph::cluster_walktrap(g.gene)$membership
    }
)
```

We counting the number of gene expression-derived subclusters in each ADT-derived parent cluster.

```{r}
data.frame(
    Cluster=names(subclusters),
    Ncells=vapply(subclusters, ncol, 0L),
    Nsub=vapply(subclusters, function(x) length(unique(x$subcluster)), 0L)
)
```

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
